Modeling the Impact of Commodity Prices and Interest Rates on the CAD/USD Exchange Rate: An Analysis of Key Drivers Using Bank Rate and BCPI

### 2024-12-15 paper/paper.qmd
remove unnecesary words

### Metal BCPI
The Metal BCPI includes the price movements of metals and minerals such as gold, silver, copper, and nickel. The prices for these commodities are collected from the London Metal Exchange (LME) and other sources, such as the Handy and Harman base price for gold and silver. These prices are used to construct an index representing the overall price changes in the metal commodities produced in Canada. Similar to the Energy BCPI, the Metal BCPI is measured in index points with the base year set to 1972.

@@

@fig-bank-agaist-exchange shows the relationship between the Weekly Bank Rate and the Weekly Average USD/CAD Exchange Rate. The data points show a positive trend, and the linear regression line suggests that as the weekly bank rate increases, the USD/CAD exchange rate also tends to rise. This indicates a positive correlation between the two variables. The upward slope of the regression line suggests that higher bank rates may have a tendency to push the exchange rate higher.

@@

@fig-energy-bcpi-and-exchange presents a dual y-axis plot displaying the Weekly Energy BCPI and the Weekly Average USD/CAD Exchange Rate. The Energy BCPI reaches its highest value around mid-2022 and gradually declines after that, showing a trend similar to the Total BCPI. The USD/CAD exchange rate fluctuates throughout the period, but there is no apparent direct relationship between the exchange rate and the Energy BCPI. This suggests that the changes in the energy-related BCPI do not correlate with fluctuations in the USD/CAD exchange rate during the observed time.

@@

@fig-metal-bcpi-and-exchange illustrates the relationship between Weekly Metal BCPI and the USD/CAD exchange rate over time. The Metal BCPI shows a clear upward trend at the beginning, peaking in March 2023, and then decreasing. By February 2024, the index almost reaches its lowest point, similar to its level in January 2021, before beginning to rise again. The USD/CAD exchange rate is plotted on the secondary y-axis, but no strong correlation is observed between the two variables, suggesting that fluctuations in the Metal BCPI may directly align with movements in the exchange rate over this period.

@@

5. **Normality of Residuals**: The residuals of the model are assumed to be normally distributed. From @fig-qq-plot, the QQ plot is approximately a straight line, which indicates that the assumption is satisfied for both of the model.

## Limitations
1. **Exclusion of Important Variables**: 
The models only includes 4 predictors (bank rate, metal BCPI, energy BCPI, and total BCPI), omitting other relevant variables that could influence the USD/CAD exchange rate. Factors such as geopolitical events, inflation rates, oil prices, and global market sentiment may lead to fluctuations in currency but are not accounted for in the current model. This exclusion could limit the model’s ability to capture the full range of drivers influencing exchange rates.

2. **Simplified Assumptions**: 
The model assumes linear relationships between predictors and the dependent variable, which may not fully reflect the complexity of real-world exchange rate dynamics. Economic relationships can often be non-linear, and there may be interactions between predictors (e.g., the combined effects of bank rates and BCPI on exchange rates) that are not captured in this linear model. Further exploration of interaction terms or more complex modeling techniques (e.g., non-linear models) could provide a more accurate representation of the underlying dynamics.

@@

- **Model Summary**: From @tbl-modelresults, Both models demonstrated strong overall fit, with R² values of 0.831 and 0.824, respectively, and adjusted R² values slightly lower due to model complexity. The inclusion of additional predictors in the First Model results in slightly higher explanatory power. The Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC) also indicate that the First Model performs better despite its increased complexity. The residual standard errors are similar, reflecting consistent performance across both models.

@@

Although the coefficients for BCPI components are relatively low, this could be attributed to the large range and average values of these indices as shown in @fig-data-sum. The **weekly energy BCPI** in particular has a broad span, with values ranging from 874.12 to over 2282.35. Thus, these results still indicate that commodity prices have a noticeable influence on the exchange rate.

@@

However, the observed results may be driven by complex underlying factors that were not captured by the model. Specifically, one potential explanation for this unexpected finding lies in **causal relationships**. In economic studies, it is often difficult to definitively establish the direction of causality between variables. Although theory suggests that higher interest rates should attract capital inflows and thus appreciate the domestic currency, in reality, it is possible that exchange rate fluctuations themselves may prompt adjustments in the **bank rate**. For example, the central bank might alter interest rates in response to exchange rate pressures to stabilize the economy.

@@

- **Economic Context: ** The study did not fully explore the broader macroeconomic context in which the exchange rate operates. Factors such as the fiscal policies, global economic conditions, and geopolitical events in USA could have significant impacts on the USA/CAD exchange rate but were not included in the model. These factors should be considered in future research to provide a more comprehensive understanding of exchange rate movements.

@@

   **Variables Collected**:  
   - **Commodity prices**: Daily or weekly prices for energy commodities (oil, natural gas), metals (gold, silver, copper), and agricultural products (wheat, corn, soybeans).
   - **Commodity index values**: GCPI and sector-specific price indices for comprehensive market coverage.

>>>>

### Metal BCPI
The Metal BCPI includes the price movements of metals and minerals such as gold, silver, copper, and nickel. The prices for these commodities are collected from the London Metal Exchange (LME) and other sources, such as the Handy and Harman base price for gold and silver. These prices are used to construct an index representing the overall price changes in the metal commodities produced in Canada. Like the Energy BCPI, the Metal BCPI is measured in index points with the base year set to 1972.

@@

@fig-bank-agaist-exchange shows the relationship between the Weekly Bank Rate and the Weekly Average USD/CAD Exchange Rate. The data points show a positive trend, and the linear regression line suggests that as the weekly bank rate increases, the USD/CAD exchange rate also tends to rise. This indicates a positive correlation between the two variables. The upward slope of the regression line suggests that higher bank rates may tend to push the exchange rate higher.

@@

@fig-energy-bcpi-and-exchange presents a dual y-axis plot displaying the Weekly Energy BCPI and the Weekly Average USD/CAD Exchange Rate. The Energy BCPI reaches its highest value around mid-2022 and gradually declines after that, showing a trend like the Total BCPI. The USD/CAD exchange rate fluctuates throughout the period, but there is no apparent direct relationship between the exchange rate and the Energy BCPI. This suggests that the changes in the energy-related BCPI do not correlate with fluctuations in the USD/CAD exchange rate during the observed time.

@@

@fig-metal-bcpi-and-exchange illustrates the relationship between Weekly Metal BCPI and the USD/CAD exchange rate over time. The Metal BCPI shows a clear upward trend at the beginning, peaking in March 2023, and then decreasing. By February 2024, the index almost reaches its lowest point, like its level in January 2021, before beginning to rise again. The USD/CAD exchange rate is plotted on the secondary y-axis, but no strong correlation is observed between the two variables, suggesting that fluctuations in the Metal BCPI may directly align with movements in the exchange rate over this period.

@@

5. **Normality of Residuals**: The residuals of the model are assumed to be normally distributed. From @fig-qq-plot, the QQ plot is approximately a straight line, which indicates that the assumption is satisfied for both models.

## Limitations
1. **Exclusion of Important Variables**: 
The models only include 4 predictors (bank rate, metal BCPI, energy BCPI, and total BCPI), omitting other relevant variables that could influence the USD/CAD exchange rate. Factors such as geopolitical events, inflation rates, oil prices, and global market sentiment may lead to fluctuations in currency but are not accounted for in the current model. This exclusion could limit the model’s ability to capture the full range of drivers influencing exchange rates.

2. **Simplified Assumptions**: 
The model assumes linear relationships between predictors and the dependent variable, which may not fully reflect the complexity of real-world exchange rate dynamics. Economic relationships can often be non-linear, and there may be interactions between predictors (e.g., the combined effects of bank rates and BCPI on exchange rates) that are not captured in this linear model. Further usage of interaction terms or more complex modeling techniques (e.g., non-linear models) could provide a more accurate representation of the underlying dynamics.

@@

- **Model Summary**: From @tbl-modelresults, both models demonstrated strong overall fit, with R² values of 0.831 and 0.824, respectively, and adjusted R² values slightly lower due to model complexity. The inclusion of additional predictors in the First Model results in slightly higher explanatory power. The Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC) also indicate that the First Model performs better despite its increased complexity. The residual standard errors are similar, reflecting consistent performance across both models.

@@

Although the coefficients for BCPI components are relatively low, this could be attributed to the large range and average values of these indices as shown in @fig-data-sum. The **weekly energy BCPI** has a broad span, with values ranging from 874.12 to over 2282.35. Thus, these results still indicate that commodity prices have a noticeable influence on the exchange rate.

@@

However, the observed results may be driven by complex underlying factors that were not captured by the model. Specifically, one potential explanation for this unexpected finding lies in **causal relationships**. In economic studies, it is often difficult to definitively establish the direction of causality between variables. Although theory suggests that higher interest rates should attract capital inflows and thus appreciate the domestic currency, it is possible that exchange rate fluctuations themselves may prompt adjustments in the **bank rate**. For example, the central bank might alter interest rates in response to exchange rate pressures to stabilize the economy.

@@

- **Economic Context: ** The study did not fully explore the broader macroeconomic context in which the exchange rate operates. Factors such as the fiscal policies, global economic conditions, and geopolitical events in USA could have significant impacts on the USA/CAD exchange rate but were not included in the model. These factors should be considered in future research to provide a better understanding of exchange rate movements.

@@

   **Variables Collected**:  
   - **Commodity prices**: Daily or weekly prices for energy commodities (oil, natural gas), metals (gold, silver, copper), and agricultural products (wheat, corn, soybeans).
   - **Commodity index values**: GCPI and sector-specific price indices for better market coverage.
